Reproducible development environments with rix

### 2024-02-07 paper.qmd
Update paper.qmd

---
title: "Reproducible development environments with rix"
format:
    jss-pdf:
        keep-tex: true
    jss-html: default
author:
  - name: Bruno Rodrigues 
    affiliations:
      - name: Ministry of Research and Higher education, Luxembourg 
        department: Department of Statistics
        address: 18, Montée de la Pétrusse
        city: Luxembourg 
        country: Luxembourg 
        postal-code: 2327
      - Journal of Statistical Software
    orcid: 0000-0002-3211-3689
    email: bruno@brodrigues.co
    url: https://www.brodrigues.co 
  - name: Philipp Baumann
    affiliations:
      - Plus Affiliation
abstract: |
  In order to create an analysis that is easily reproducible,
  it is not enough to write clean code and document it well. One
  must also make sure to clear list all the dependencies of the analysis
  and ideally provide an easy way to install them.
  There are several tools that can be used to list dependencies and to
  make them easily installable by someone that wishes to reproduce a study,
  such as [Docker]{.proglang}, a containerization solution. This paper will
  present another tool that can be used to make projects reproducible,
  the Nix package manager, and an [R]{.proglang} package called [rix]{.pkg} that
  makes using Nix much simpler for users of [R]{.proglang}.

@@

The next step in reaching the gold standard of reproducibility would be to not
only install the right packages used for the analysis, but also the right
version of [R]{.proglang}. Of course, it would be possible to install the right
version manually, but here too, there are tools that simplify the process. One
such tool is [rig]{.proglang} by the R infrastructure team [-@rlib2023].
[rig]{.proglang} makes it easy to install different versions of [R]{.proglang},
so one could use it to install the right version to reproduce an analysis, and
then use one of the listed packages above to install the right library of
[R]{.proglang} packages. However, this involves many manual steps and is thus
error prone. It is also time-consuming to do so, especially if one wants to
reproduce many studies.

The final tool one could use to reach the gold standard of reproducibility is to
package the right version of [R]{.proglang} and [R]{.proglang} packages inside a
[Docker]{.proglang} image. [Docker]{.proglang} is a containerisation tool: using
[Docker]{.proglang} it is possible to package some *data product* with its
dependencies into an image. A statistical analysis, from the simplest to the
most complex one, can be seen as such a data product that requires many software
dependencies. *Dockerizing* an analysis consists in first building an image: in
the build step, the dependencies of the analysis have to be installed and this
can be achieved using the tools mentioned above. The scripts to run the analysis
and the data are also added to the image at build time. These steps are written
down in a simple text file called a `Dockerfile`. This `Dockerfile` is the
definition of the [Docker]{.prolgang} image.

It is possible to then execute the code of a study inside a so-called container,
which is an instance of an image. Executing, or running, these containers is
done non-interactively, and so it is possible to instantiate the correct
environment to run a study in a single command.

The advantage of using [Docker]{.proglang} is not only about having the right
version of [R]{.proglang} and [R]{.proglang} packages bundled alongside the
analysis scripts: [Docker]{.proglang} images are actually almost complete Linux
operating systems, which means that system-level dependecies also get bundled
into the image. This is especially useful for packages that rely on external
libraries to run, such as for [nloptr]{.pkg}, as these external libraries are
thus also made available to future reproducers of the study within the
[Docker]{.proglang} image. The image can also be shared.

The Rocker project initiated by @boettiger2017 provides many pre-built
[Docker]{.proglang} to the [R]{.proglang} community of users, already containing
specific [R]{.proglang} versions and even packages. These images can be used as
bases to build other images containing statistical analyses more easily than
starting from a bare-bones image.

While [Docker]{.proglang} offers a great solution to the reproducibility
problem, it can be awkward to work interactively from a running container:
containers can be changed at run-time, but any change will be lost once the
container is stopped. Also, while it is possible to launch graphical programs
from a container, this is generally not easy to set up and only works, as far as
we know, only on Linux operating systems. In most cases, users of
[Docker]{.proglang} will instead use the container to run a web server and the
web based application of a popular integrated development environment (IDE). So
for instance, it would be possible to launch the web interface of the RStudio
IDE, but not the operating system client of RStudio, which is usually what an R
programmer would use (or VS Code or Emacs...). A common approach seems to be to
work on a study interactively using a standard installation of R and R packages,
and after the study is done, take some time to write the required `Dockerfile`
to provide an easy way for future replicators to reproduce the results of the
study. But this approach does not solve the problem of deploying the same
environment on several machines, and so ideally authors would develop using
[Docker]{.proglang}. Another challenge of using [Docker]{.proglang} is that,
as stated already, [Docker]{.proglang} images are actually almost complete
Linux operating systems, so experience with Linux is highly recommended if
one wishes to properly write a `Dockerfile`.

A final, and optional, step towards the gold standard is to use a build
automation tool such as [Make]{.proglang} to run the whole analysis when the
container is executed, instead of executing scripts. Build automation tools make
it easier to run arbitrary code in a series of well-defined steps.
[R]{.proglang} programmers can use the [targets]{.pkg} by @landau2021 as a build
automation tool.

@@

However, reaching this gold standard is quite costly: one needs to learn a tool
to deal with package versions, then [Docker]{.proglang} for all the other
software, including the programming language that was used. Ideally, a build
automation tool such as [Make]{.proglang} should be added to the list. It should
also be noted that these requirements are not limited to the [R]{.proglang}
programming language. A very similar approach should be taken if one uses
[Python]{.proglang} for statistical analysis instead. Furthermore, the previous
discussion also implies that for an analysis to actually be reproducible at all,
it must be made using free and open source software, as proprietary software
cannot be freely shared with others that wish to reproduce a study, and there's
no guarantee that others will have access to the proprietary software used.

As an alternative to the above approach, we will present the [Nix]{.proglang}
package manager, which is available for all major operating systems and focuses
on installing and building software in a reproducible manner. This means that
this single tool can replace not only [Docker]{.proglang} but also [renv]{.pkg}
and even a build automation tool such as [Make]{.proglang}. To make
[Nix]{.proglang} more accessible to [R]{.proglang} programmers, we developed the
[rix]{.pkg} package which we will present in this article.

@@

[Nix]{.proglang} is a package manager that can be used to install and build
software in a completely reproducible manner. As of writing, it contains more
than 80.000 packages, including the entirety of CRAN and Bioconductor. This
means that using [Nix]{.proglang}, it is possible to install not only
[R]{.proglang}, but also all the packages required for a project. The reason why
one should use [Nix]{.proglang} to install [R]{.proglang} packages, and not use
the usual, built-in, `install.packages()` [R]{.proglang} function instead, is
that when installing a package with [Nix]{.proglang}, [Nix]{.proglang} makes
sure to install every dependency of every package, whether this dependency is
another [R]{.proglang} package or a system-level dependency.

For example, the [xlsx]{.pkg} [R]{.proglang} package requires the
[Java]{.proglang} programming language to be installed to successfully install.
Depending on what system one tries to install [xlsx]{.pkg}, installing
[Java]{.proglang} might not be easy, nor even possible. But with
[Nix]{.proglang}, it suffices to declare that the [xlsx]{.pkg} package is needed
for the project, and [Nix]{.proglang} figures out automatically that
[Java]{.proglang} is required and installs and configures it. It all just
happens without any required intervention from the user.

But where do these packages come from? When installing a package using
[Nix]{.proglang}, an expression written in the [Nix]{.proglang} programming
language gets downloaded from the `nixpkgs` Github repository and evaluated.
This expression contains a so-called *derivation*. A derivation defines a build:
its dependencies, commands to build and install the package in question, and
then an output. Most of the time, a derivation downloads source code, builds the
software from the source and then outputs a compiled binary. Derivations are
extremely flexible, and by writing one's own, it is possible to define and build
an environment for a project in a reproducible manner.

But why is installing software with [Nix]{.proglang} reproducible? Because the
whole set of [Nix]{.proglang} expressions is hosted on Github, it is possible to
use a specific commit of `nixpkgs` to ensure reproducibility of our project.
Using a specific commit of `nixpkgs` is called *pinning a revision (or commit)*.
Pinning a revision ensures that every package that Nix installs will always be
at exactly the same versions, regardless of when in the future the packages get
built. This is because the [Nix]{.proglang} expressions that get downloaded when
pinning are the expressions as they were at that commit of `nixpkgs`. This is
not the only reason [Nix]{.proglang} is very useful for reproducibility, but the
[Nix]{.proglang} package manager itself was designed with reproducibility in
mind. [Nix]{.proglang} is functional package manager, meaning that it uses
concepts from functional programming to build software: no-side effects or
global variables are allowed, and the output is always the same if given the
same inputs, regardless of where or when a package gets built.
We refer to @dolstra2004nix for a more technical discussion on [Nix]{.proglang}.

@@

@@

[Nix]{.proglang} expressions are written in the [Nix]{.proglang} programming language,
which is a purely functional programming language. Here is a simple expression
that builds a shell environment containing version 4.3.1 of [R]{.proglang}:

@@

To define a variable in the [Nix]{.proglang}, the `let` keyword is used, so the
first lines of this expression define a variable called `pkgs` and
`system_pkgs`. `pkgs` is defined as the set of packages to be used in this
derivation. In this example, the package expressions are going to be pulled from
the official `nixpkgs` Github repository, and in particular the commit starting
with `976fa336` is going to be used. It should be noted that is also possible to
use a personal fork of `nixpkgs` instead of the official repository.
`system_pkgs` lists the packages that we wish to install and use with this
environment, in this example only the [R]{.proglang} programming language will
be installed, alongside all the required dependencies to actually install and
run [R]{.proglang}. Finally, the `mkShell` function is used to build a shell
containing the set of packages defined by the variable `system_packages` and its
`shellHook` is set to `R --vanilla`: this means that when entering the shell,
[R]{.proglang} will be started with the `--vanilla` flag (thus ignoring any
startup options).

>>>>

---
title: "Reproducible development environments with rix"
format:
    jss-pdf:
        keep-tex: true
    jss-html: default
author:
  - name: Bruno Rodrigues 
    affiliations:
      - name: Ministry of Research and Higher education, Luxembourg 
        department: Department of Statistics
        address: 18, Montée de la Pétrusse
        city: Luxembourg 
        country: Luxembourg 
        postal-code: 2327
      - Journal of Statistical Software
    orcid: 0000-0002-3211-3689
    email: bruno@brodrigues.co
    url: https://www.brodrigues.co 
  - name: Philipp Baumann
    affiliations:
      - Plus Affiliation
abstract: |
  In order to create an analysis that is easily reproducible,
  it is not enough to write clean code and document it well. One
  must also make sure to clearly list all the dependencies of the analysis
  and ideally provide an easy way to install them.
  There are several tools that can be used to list dependencies and to
  make them easily installable by someone that wishes to reproduce a study,
  such as [Docker]{.proglang}, a containerization solution. This paper will
  present another tool that can be used to make projects reproducible,
  the Nix package manager, and an [R]{.proglang} package called [rix]{.pkg} that
  makes using Nix much simpler for users of [R]{.proglang}.

@@

The next step in reaching the gold standard of reproducibility would be to not
only install the right packages used for the analysis, but also the right
version of [R]{.proglang}. Of course, it would be possible to install the right
version manually, but here too, there are tools that simplify the process. One
such tool is [rig]{.pkg} by the R infrastructure team [-@rlib2023].
[rig]{.pkg} makes it easy to install different versions of [R]{.proglang},
so one could use it to install the right version to reproduce an analysis, and
then use one of the listed packages above to install the right library of
[R]{.proglang} packages. However, this involves many manual steps and is thus
error prone and time-consuming.

The final tool one could use to reach the gold standard of reproducibility is to
package the right version of [R]{.proglang} and [R]{.proglang} packages inside a
[Docker]{.pkg} image. [Docker]{.pkg} is a containerisation tool: using
[Docker]{.pkg} it is possible to package some *data product* with its
dependencies into an image. A statistical analysis, from the simplest to the
most complex one, can be seen as such a data product that requires many software
dependencies. *Dockerizing* an analysis consists in first building an image: in
the build step, the dependencies of the analysis have to be installed and this
can be achieved using the tools mentioned above. The scripts to run the analysis
and the data are also added to the image at build time. These steps are written
down in a simple text file called a `Dockerfile`. This `Dockerfile` is the
definition of the [Docker]{.pkg} image.

It is possible to then execute the code of a study inside a so-called container,
which is a running instance of an image. Executing, or running, these containers is
done non-interactively, and so it is possible to instantiate the correct
environment to run a study in a single command.

The advantage of using [Docker]{.pkg} is not only about having the right
version of [R]{.proglang} and [R]{.proglang} packages bundled alongside the
analysis scripts: [Docker]{.pkg} images are actually almost complete Linux
operating systems, which means that system-level dependencies also get bundled
into the image. This is especially useful for packages that rely on external
libraries to run, such as for [nloptr]{.pkg}, as these external libraries are
thus also made available to future reproducers of the study within the
[Docker]{.pkg} image. The image can also be shared.

The Rocker project initiated by @boettiger2017 provides many pre-built
[Docker]{.pkg} to the [R]{.proglang} community of users, already containing
specific [R]{.proglang} versions and even packages. These images can be used as
bases to build other images containing statistical analyses more easily than
starting from a bare-bones image.

While [Docker]{.pkgs} offers a great solution to the reproducibility
problem, it can be awkward to work interactively from a running container:
containers can be changed at run-time, but any change will be lost once the
container is stopped. Also, while it is possible to launch graphical programs
from a container, this is generally not easy to set up and only works, as far as
we know, only on Linux operating systems. In most cases, users of
[Docker]{.pkg} will instead use the container to run a web server and the
web based application of a popular integrated development environment (IDE).
A common approach seems to be to
work on a study interactively using a standard installation of R and R packages,
and after the study is done, take some time to write the required `Dockerfile`
to provide an easy way for future replicators to reproduce the results of the
study. But this approach does not solve the problem of deploying the same
environment on several machines, and so ideally authors would develop using
a containerized development environment. Another challenge of using [Docker]{.pkg} is that,
as stated already, [Docker]{.pkg} images are actually almost complete
Linux operating systems, so experience with Linux is highly recommended if
one wishes to properly write a `Dockerfile`. Finally, one also needs to be careful
and write the `Dockerfile` in such a way that the build process will always result
in the same image. This is only possible if a specific digest of a base
image is used instead of a tagged version, but in practice, this is rarely done.

A final, and optional, step towards the gold standard is to use a build
automation tool such as [Make]{.pkg} to run the whole analysis when the
container is executed, instead of executing scripts. Build automation tools make
it easier to run arbitrary code in a series of well-defined steps.
[R]{.proglang} programmers can use the [targets]{.pkg} by @landau2021 as a build
automation tool.

@@

However, reaching this gold standard is quite costly: one needs to learn a tool
to deal with package versions, then [Docker]{.pkg} for all the other
software, including the programming language that was used. Ideally, a build
automation tool such as [Make]{.proglang} should be added to the list. It should
also be noted that these requirements are not limited to the [R]{.proglang}
programming language. A very similar approach should be taken if one uses
[Python]{.proglang} for statistical analysis instead. Furthermore, the previous
discussion also implies that for an analysis to actually be reproducible at all,
it must be made using free and open source software, as proprietary software
cannot be freely shared with others that wish to reproduce a study, and there's
no guarantee that others will have access to the proprietary software used.

As an alternative to the above approach, we will present the [Nix]{.proglang}
package manager, which is available for all major operating systems and focuses
on installing and building software in a reproducible manner. This means that
this single tool can replace not only [Docker]{.pkg} but also [renv]{.pkg}
and even a build automation tool such as [Make]{.pkg}. To make
[Nix]{.proglang} more accessible to [R]{.proglang} programmers, we developed the
[rix]{.pkg} package which we will present in this article.

@@

[Nix]{.pkg} is a package manager that can be used to install and build
software in a completely reproducible manner. As of writing, it contains more
than 80.000 packages, including almost the entirety of CRAN and Bioconductor. This
means that using [Nix]{.proglang}, it is possible to install not only
[R]{.proglang}, but also all the packages required for a project. The reason why
one should use [Nix]{.proglang} to install [R]{.proglang} packages, and not use
the usual, built-in, `install.packages()` [R]{.proglang} function instead, is
that when installing a package with [Nix]{.pkg}, [Nix]{.pkg} makes
sure to install every dependency of every package, whether this dependency is
another [R]{.proglang} package or a system-level dependency.

For example, the [xlsx]{.pkg} [R]{.proglang} package requires the
[Java]{.proglang} programming language to be installed to successfully install.
Depending on what system one tries to install [xlsx]{.pkg}, installing
[Java]{.proglang} might not be easy, nor even possible. But with
[Nix]{.pkg}, it suffices to declare that the [xlsx]{.pkg} package is needed
for the project, and [Nix]{.proglang} figures out automatically that
[Java]{.proglang} is required and installs and configures it. It all just
happens without any required intervention from the user. In this context, 
[xlsx]{.pkg} can be refered to as a component closure, and quoting @dolstra2004nix:

> The idea is to always deploy component closures: if we deploy a
> component, then we must also deploy its dependencies, their dependencies, and so on. That is, we must
> always deploy a set of components that is closed under
> the ‘‘depends on’’ relation. Since closures are selfcontained, they are the units of complete software
> deployment. After all, if a set of components is not
> closed, it is not safe to deploy, since using them might
> cause other components to be referenced that are missing on the target system.

But how does [Nix]{.pkg} achieve this? Where do these packages, or closures, come from? When installing a package using
[Nix]{.pkg}, an expression written in the [Nix]{.proglang} programming
language gets downloaded from the `nixpkgs` Github repository and evaluated.
This expression contains a so-called *derivation*. A derivation defines a build:
its dependencies, commands to build and install the package in question, and
then an output. Most of the time, a derivation downloads source code, builds the
software from the source and then outputs a compiled binary. Derivations are
extremely flexible, and by writing one's own, it is possible to define and build
an environment for a project in a reproducible manner. The dependencies of these
derivations are also defined in other expressions, which will also get installed
if needed.

But why is installing software with [Nix]{.proglang} reproducible? Because the
whole set of [Nix]{.pkg} expressions is hosted on Github, it is possible to
use a specific commit of `nixpkgs` to ensure reproducibility of our project.
Using a specific commit of `nixpkgs` is called *pinning a revision (or commit)*.
Pinning a revision ensures that every package that Nix installs will always be
at exactly the same versions, regardless of when in the future the packages get
built. This is because the [Nix]{.pkgs} expressions that get downloaded when
pinning are the expressions as they were at that commit of `nixpkgs`. This is
not the only reason [Nix]{.pkg} is very useful for reproducibility, but the
[Nix]{.pkg} package manager itself was designed with reproducibility in
mind. [Nix]{.pkg} is functional package manager, meaning that it uses
concepts from functional programming to build software: no-side effects or
global variables are allowed, and the output is always the same if given the
same inputs, regardless of where or when a package gets built. [Nix]{.pkg} 
also allows for multiple versions (or variants in the [Nix]{.pkg} jargon)
to be installed at the same time on the same system. Each variant of a package
has a unique identifier, so that it is possible to have two versions of [R]{.proglang}
installed and use the correct one where adequate.
We refer to @dolstra2004nix for a more technical discussion on [Nix]{.pkg}.

@@

[Nix]{.pkg} has a fairly high entry cost though. [Nix]{.pkg} is a complex piece of software that
comes with its own programming language, which is also called [Nix]{.proglang}. Its purpose is 
to solve a complex problem: defining instructions on how to build software packages 
and manage configurations in a declarative way. This makes sure that software gets 
installed in fully reproducible manner, on any operating system or hardware. To increase
[Nix]{.pkg}’s adoption, we wrote the [rix]{.pkg} package.

@@

[Nix]{.pkg} expressions are written in the [Nix]{.proglang} programming language,
which is a purely functional programming language. Here is a simple expression
that builds a shell environment containing version 4.3.1 of [R]{.proglang}:

@@

To define a variable in the [Nix]{.proglang} language, the `let` keyword is used, so the
first lines of this expression define a variable called `pkgs` and
`system_pkgs`. `pkgs` is defined as the set of packages to be used in this
derivation. In this example, the package expressions are going to be pulled from
the official `nixpkgs` Github repository, and in particular the commit starting
with `976fa336` is going to be used. It should be noted that is also possible to
use a personal fork of `nixpkgs` instead of the official repository.
`system_pkgs` lists the packages that we wish to install and use with this
environment, in this example only the [R]{.proglang} programming language will
be installed, alongside all the required dependencies to actually install and
run [R]{.proglang}. Finally, the `mkShell` function is used to build a shell
containing the set of packages defined by the variable `system_packages` and its
`shellHook` is set to `R --vanilla`: this means that when entering the shell,
[R]{.proglang} will be started with the `--vanilla` flag (thus ignoring any
startup options).
